From heartbeat to data -- Using wearable fitness trackers as an affordable approach to assess teacher stress

### 2024-09-27 02_paper/02_study/01_fitbit paper/Submission/FitBit_submission_template.tex
nochmal formatiert

\documentclass[preprint, 3p,
authoryear]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

@@

\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{booktabs}
\usepackage{array}
\usepackage{caption}
\usepackage{tablefootnote}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{pdflscape}
\usepackage{longtable}
\usepackage{geometry}
\usepackage{caption}

@@

  \title{From heartbeat to data -- Using wearable fitness trackers as an
affordable approach to assess teacher stress}
    \author[Leipzig University]{Mandy Klatt%
  \corref{cor1}%
  \fnref{1}}
   \ead{mandy.klatt@uni-leipzig.de} 
    \author[University of Greifswald]{Peer Keßler%
  %
  \fnref{2}}
   \ead{peer.kessler@uni-greifswald.de} 
    \author[Max Planck Institute for Evolutionary Anthropology]{Gregor
Kachel%
  %
  \fnref{3}}
   \ead{gregor@example.com} 
    \author[Leipzig University]{Christin Lotz%
  %
  \fnref{4}}
   \ead{christin.lotz@uni-leipzig.de} 
    \author[Leipzig University]{Anne Deiglmayr%
  %
  \fnref{5}}
   \ead{anne.deiglmayr@uni-leipzig.de} 
      \affiliation[Leipzig University]{
    organization={Empirische Schul- und
Unterrichtsforschung},addressline={Marschnerstr.
29},city={Leipzig},postcode={4109},state={Sachsen},country={Germany},}
    \affiliation[Max Planck Institute for Evolutionary Anthropology]{
    organization={Department},addressline={Deutscher Pl.
6},city={Leipzig},postcode={4103},state={Sachsen},country={Germany},}
    \affiliation[University of Greifswald]{
    organization={Institute of Psychology},addressline={Robert-Blum-Str.
13},city={Greifswald},postcode={17489},state={Mecklenburg-Vorpommern},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{Conceptualization, Methodology, Formal analysis,
Investigation, Data Curation, Writing - Original Draft}
    \fntext[2]{Conceptualization, Methodology, Writing - Review \&
Editing, Supervision}
    \fntext[3]{Formal analysis, Data Curation, Writing - Review \&
Editing}
    \fntext[4]{Conceptualization, Methodology, Writing - Review \&
Editing}
    \fntext[5]{Conzeptualization, Methodology, Writing - Review \&
Editing, Supervision}

@@

Figure \ref{fig.1} shows, in a simplified way, how classroom events
affect teachers' stress level, according to the adaptation of the
Lazarus model to teacher stress proposed by \citet{van2006stress}: When
potential stressors (e.g., classroom disruptions) occur during teaching
(1), teachers intuitively judge how disruptive the event is in a primary
appraisal (2). If potential stressors are judged as threatening, i.e.,
as actual stressors (3), teachers consider whether they have sufficient
resources for coping with the stressors (4). Teachers utilize these
resources in trying to cope with the stressors, e.g., by employing
classroom management strategies (5). In cases when coping fails, stress
ensues, often accompanied by physiological reactions like increased HR
(6). As part of the coping process, and dependent on its outcomes,
teachers re-evaluate the stressor (7).

As shown in Figure \ref{fig.1}, teachers' primary and secondary
appraisal as well as coping attempts are influenced by professional
experience. As professional experience grows, teachers develop cognitive
scripts for managing classroom events, resulting in more complex and
problem-focused classroom management skills \citep{wolff2021classroom},
and thus more effective coping and less stress. Empirically, classroom
management skills and problem-focused coping styles are linked to fewer
instances of emotional exhaustion
\citep{maslach2001job, clunies2008self}. Novices in the teaching
profession, on the other hand, face considerable stress and often feel
overwhelmed by the demands of teaching
\citep{ophardt2017klassenmanagement, wolff2015keeping, klusmann2012berufliche}
with many leaving the profession within the first five years
\citep{ingersoll2003}. Accordingly, when resources are lacking and
coping fails, negative consequences for health (e.g., burnout) and for
work (e.g., high turnover rates) can arise
\citep{jalongo2006, unterbrink2007, aloe2014multivariate}, highlighting
the importance of professional expertise for managing teacher stress
\citep{fisher2011}.

\captionsetup[figure]{
    labelfont=bf,       % Figure label bold (not italic)
    labelsep=newline,      % Period after Figure label
    textfont=it,          % Italic font for the caption text
    font=small,           % Smaller font size for APA style
    justification=justified, % Left-justified captions, as per APA guidelines
    singlelinecheck=false % Avoid centering if the caption is single-line
}

\captionsetup[table]{
    labelfont=bf,        % Figure label bold (not italic)
    labelsep=newline,      % Period after Figure label
    textfont=it,          % Italic font for the caption text
    font=small,           % Smaller font size for APA style
    justification=justified, % Left-justified captions, as per APA guidelines
    singlelinecheck=false % Avoid centering if the caption is single-line
}

\begin{figure}[htbp]
  \caption{\\
  A model of teacher stress (adapted from van Dick 2006, p.37, modified by the authors)}
  \centering
  \includegraphics[width=1\textwidth]{images/Model_Teacher_Stress_adapted_new.pdf}
  \label{fig.1}
\end{figure}

@@

In addition, we selected five representative 10-minute intervals from
the five phases of the lab study (see Figure \ref{fig.2}) in order to
test the hypotheses that, regarding HR levels, teachers' HR would be the
highest during the micro-teaching unit, compared to all other phases
(\textbf{Hypothesis 1a}), and, regarding HR slopes, that teachers' HR
would increase while they were preparing for teaching (pre-teaching
interval), but decrease in all of the following intervals, i.e.~when
they were habituating to and recovering from the stressful
micro-teaching unit (\textbf{Hypothesis 1b}).

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  We further explored whether teaching experience made a difference in
  how teachers' HR reacted to the classroom disruptions. We expected
  more experienced teachers to be less stressed by the classroom events
  (\textbf{Hypothesis 2a}). In addition, we examined the relations
  between teachers' subjective appraisals of the classroom events
  (specifically, the disruptiveness of the events, and their confidence
  in dealing with them) and teachers' HR level, beyond the explanatory
  power of teaching experience. We expected higher HR levels for
  teachers who felt more disrupted, regardless of their teaching
  experience (\textbf{Hypothesis 2b}), and lower HR levels for teachers
  who felt more confident, regardless of teaching experience
  (\textbf{Hypothesis 2c}). We hypothesized that each of the three
  predictors (\emph{teaching experience, disruption appraisal,
  confidence appraisal}) uniquely contributed to explaining variance in
  teachers' HR levels (\textbf{Hypothesis 2d}). In addition, we
  exploratively ran analogous analyses for the \emph{changes} in HR
  (i.e., slopes).
\end{enumerate}

@@

Each participant came to the lab for a period of approximately two hours
in total, and each participant underwent the same phases (see Figure
\ref{fig.2}): In the \emph{pre-teaching phase}, the experimenter
welcomed the participants and helped them put on the fitness tracker.
This was followed by a warm-up session to familiarize the participants
with the laboratory setting and the class. This phase took about 10-15
minutes and participants spent this time mostly standing or slowly
walking around. During the \emph{teaching phase}, the participants held
their self-prepared micro-teaching unit to a class of three trained
actors who performed nine, potentially disruptive, classroom events
(e.g., chatting with a neighbor, heckling, looking at the phone; see
Table \ref{tab_a1} in the supplementary material for an overview and
categorization of all events; and Figure \ref{fig.a2} in the
supplementary material for a depiction of the laboratory setting of the
micro-teaching unit). The topic and class level of the teaching unit
could be freely chosen by the teachers with the only requirement that
the unit had to be an introductory lesson, and had to consist of
supervised individual work and / or frontal teaching. The micro-teaching
unit lasted about 15-20 minutes. Participants spent this time mostly
standing or slowly walking around. While teaching, participants wore
eye-tracking glasses, and their lesson was video-recorded. After having
completed the micro-teaching unit, in the \emph{post-teaching phase},
participants filled in questionnaires for approximately 10-15 minutes: a
brief computer-based survey of sociodemographic

\begin{wrapfigure}[36]{r}{0.5\textwidth}
  \centering
  \caption{\\
  Procedure of the two-hour-long study consisting of five phases with five representative 10-minute intervals}
  \includegraphics[width=0.5\textwidth]{images/Timeline_smaller.pdf}
  \label{fig.2}
\end{wrapfigure}

\noindent data (e.g., teaching experience, gender, studied school type,
studied school subjects, extracurricular teaching activities), and a
short knowledge test that was irrelevant to the present study. In the
\emph{interview phase}, participants engaged in a Stimulated Recall
Interview (SRI). During the SRI, participants sat in front of a computer
monitor and watched the video of their own lesson from the ego
perspective, as recorded through the eye-tracking glasses. The
experimenter stopped the video each time one of the nine classroom
events happened, and asked five open-ended, and three rating questions
per event. Two of the rating questions are relevant to the present
study: the disruption and the confidence appraisal ratings (see
Measures). The interview lasted about 45-60 minutes. Finally, in the
\emph{end phase}, participants filled in another questionnaire
irrelevant to the present study, which lasted about 10-15 minutes.

@@

Since we aimed to keep measurement intervals comparable between study
phases, we aggregated HR over a representative 10-minute interval within
each phase (cf.~Figure \ref{fig.2}). Previous research has indicated
that 10-minute intervals are a useful duration for analyzing PPG data
\citep{lu2008can}. The intervals were selected based on the following
rules: The \emph{pre-teaching interval} (\(I_1\)) comprised the first 10
minutes after the fitness tracker had been put on. The \emph{teaching
interval} (\(I_2\)) started two minutes after the lesson had started.
This interval was of the highest relevance to our study. We explicitly
chose an early 10-minute interval within the teaching phase, as previous
studies revealed that the beginning of a lesson is most demanding and
potentially stressful with regards to teacher-student interaction
\citep{donker2018, claessens2017positive}. The \emph{post-teaching
interval} (\(I_3\)) started immediately after the end of the teaching
unit. The \emph{interview interval} (\(I_4\)) was defined as the mid-10
minutes between the end of the teaching unit and the time point when the
fitness tracker was taken off. All participants were being interviewed
during this interval. The end interval (\(I_5\)) comprised the last 10
minutes before the fitness tracker was taken off.

@@

To enable visual inspection of HR trends, we displayed smoothed teacher
HR over the course of the recording.\footnote{The curve was smoothed
  using the geom\_smooth() function from the ggplot2 package in R
  \citep{ggplot2} based on the smoothing method LOESS (Locally Estimated
  Scatterplot Smoothing). This method fits a polynomial surface
  determined by one or more numerical predictors, using local fitting.}
We visually compared unstandardized and standardized HR trends over the
two-hour recording period.\footnote{Note that the study exceeded the
  planned duration of two hours for a few participants. To avoid
  distortions when mapping the HR over the course of the study (see
  Figure \ref{fig.3}), the endpoint was set at two hours for all
  participants, even though data from later time points was used in the
  \emph{end interval} for a few participants.} For all further analyses,
we used standardized rather than unstandardized HR values.

@@

For testing Hypothesis 1b, concerning HR changes within each interval,
we first conducted a linear estimation of the increase or decrease in
standardized HR values over time for each participant. To this end, we
used fixed intercept fixed slope regression models
\citep{gelman2006data} for each interval to estimate intercepts and
linear slopes for each individual, which were then averaged across
individuals.\footnote{Although this procedure does not account for
  nonmonotonic progressions in individual HR, a graphical evaluation
  revealed that the linear estimates corresponded well to the majority
  of the cases (see XX in the supplementary material).} We tested
Hypothesis 1b based on the unstandardized estimates of mean slopes (one
estimate per participant per interval).

@@

Means, standard deviations, and range of teachers' unstandardized and
standardized HR for the entire study period, and for the five intervals,
are shown in Table \ref{tab_1}. Figure \ref{fig.3} displays the
unstandardized and standardized HR trends, respectively, over the course
of the entire study period. HR initially increased, peaked, and then
decreased, with the unstandardized and standardized HR graphs showing
high similarity. Thus, for all further analyses, we used participants'
standardized HR values.

Figure \ref{fig.4} shows the distribution of teachers' mean standardized
HR for the five intervals. Repeated measures ANOVA revealed significant
differences in mean standardized HR between intervals,
\(F(4, 400) = 260.62\), \(p < .05\), \(d = 1.60\) (large effect).
Planned contrasts indicated that, as hypothesized (Hypothesis 1a), mean
standardized HR was significantly higher in the \emph{teaching interval}
(\(I_2\)) than in all other intervals, specifically, the
\emph{pre-teaching interval} (\(I_1\); \(t(400) = -10.08\), \(p < .05\),
\(d = 1.034\); large effect), the post-teaching interval (\(I_3\);
\(t(400) = -6.94\), \(p < .05\), \(d = 1.37\); large effect), the
interview interval (\(I_4\); \(t(400) = 15.00\), \(p < .05\),
\(d = 3.29\); large effect), and the end interval (\(I_5\));
\(t(400) = 22.54\), \(p < .05\), \(d = 4.64\); large effect).

Next, we examined HR changes (i.e., mean slopes) within each interval to
test the hypothesis that HR increased during the \emph{pre-teaching
phase} and decreased during all other phases (Hypothesis 1b). The mean
intercepts and mean slopes, complemented by their standard deviations
for each interval, are shown in Table 2. The mean slope of the
\emph{pre-teaching interval} (\(I_1\)) was significantly positive,
indicating an increase in HR, as hypothesized. Further, the mean slopes
of the \emph{teaching interval} (\(I_2\)), \emph{post-teaching interval}
(\(I_3\)), and \emph{interview interval} (\(I_4\)) were significantly
negative, indicating a decrease in HR. For the last interval, the
\emph{end interval} (\(I_5\)), the mean slope was negative, but did not
differ significantly from zero.

@@

\begin{figure}[H]
  \centering
  \caption{\\
  Overall course of the HR with the unstandardized HR in BPM shown in Fig. 3a. and the z-standardized HR shown in Fig. 3b. for the planned 2-hour study.}
  \includegraphics[width=1\textwidth]{plots_publication/loess_plot_std_unstd_new.pdf}
  \label{fig.3}
\end{figure}

@@

\begin{table}[ht]
    \centering
    \caption{\\
    Analysis (\textit{M, SD, p}-values) for the mean intercepts and the mean slopes for the five intervals.}
    \begin{tabularx}{\textwidth}{lXXXXXX}
        \toprule
        Interval  & \multicolumn{2}{c}{\textit{M} (\textit{SD})} & \multicolumn{2}{c}{$p$} \\ & Intercept & Slope & Intercept & Slope \\
        \midrule
        (1) Pre-teaching interval & 0.052 (0.820) & 0.085 (0.133) & .57 & $< .05$ \\
        (2) Teaching interval & 1.025 (0.690) & -0.039 (0.108) & $< .05$ & $< .05$ \\
        (3) Post-teaching interval & 0.549 (0.547) & -0.060 (0.101) & $< .05$ & $< .05$ \\
        (4) Interview interval & -0.617 (0.614) & -0.022 (0.070) & $< .05$ & $< .05$ \\
        (5) End interval & -1.004 (0.500) & -0.012 (0.074) & $< .05$ & .14 \\
        \bottomrule
    \end{tabularx}
    \label{tab_2}
\end{table}

\begin{figure}[H]
  \centering
  \caption{\\
  Distribution of the standardized heart rate means in the five intervals}
  \includegraphics[width=1\textwidth]{plots_publication/box_plot.pdf}
  \label{fig.4}
\end{figure}

@@

Table \ref{tab_3} shows the raw correlations among mean standardized
HR/mean slopes (see Table \ref{tab_2} for means and standard
deviations), teaching experience (\(M = 5.64\), \(SD = 9.46\)),
disruption appraisal (\(M = 5.19\), \(SD = 2.87\)), and confidence
appraisal (\(M = 7.81\), \(SD = 1.97\)). With a few notable exceptions,
correlations with HR measures were mostly very small and statistically
non-significant (Table 3). Correlations between teaching experience and
appraisals (not shown in Table \ref{tab_3}) were substantial: more
experienced teachers gave lower disruption appraisals (\(r = -.36\)),
and higher confidence appraisals (\(r = .44\)). Moreover, the two
appraisal variables were negatively correlated (\(r = -.37\)).

Table \ref{tab_4} shows the results of the regression analyses. Teaching
experience significantly predicted mean standardized HR only in the
\emph{interview interval} (Table \ref{tab_4}, \emph{interview interval},
Model 1), indicating a higher mean standardized HR for teachers with
more teaching experience. This relationship is, in fact, in the opposite
direction as predicted by Hypothesis 2a. Neither adding disruption
appraisal (Hypothesis 2b) nor adding confidence appraisal (Hypothesis
2c) increased the amount of explained variance to a statistically
significant extent.

When considering the effects of the three predictors in concert
(Hypothesis 2d), mean standardized HR was significantly predicted only
by disruption appraisal, and only in the \emph{post-teaching interval}
(Table \ref{tab_4}, \emph{post-teaching interval}, Model 4), indicating
a higher mean standardized HR for teachers who felt more disrupted by
the classroom events, when controlling for the other variables.

@@

Concerning the explorative investigation of the effects of teaching
experience and subjective appraisals on \emph{changes} (i.e., mean
slopes) in teachers' HR, teaching experience significantly predicted the
mean slope in the \emph{pre-teaching interval} (Table 4, Pre-teaching
interval, Model 1), indicating a less steep HR increase in teachers with
more teaching experience. For all other intervals, no variable had
significant predictive value.

>>>>

\documentclass[preprint,
3p]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

@@

\setcitestyle{sort&compress}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{booktabs}
\usepackage{array}
\usepackage{caption}
\usepackage{tablefootnote}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{pdflscape}
\usepackage{longtable}
\usepackage{geometry}
\usepackage{caption}

@@

  \title{From heartbeat to data -- Using wearable fitness trackers as an
affordable approach to assess teacher stress}
    \author[Leipzig University]{Mandy Klatt%
  \corref{cor1}%
  \fnref{1}}
   \ead{mandy.klatt@uni-leipzig.de} 
    \author[University of Greifswald]{Peer Keßler%
  %
  \fnref{2}}
   \ead{peer.kessler@uni-greifswald.de} 
    \author[Max Planck Institute for Evolutionary Anthropology]{Gregor
Kachel%
  %
  \fnref{3}}
   \ead{gregor@example.com} 
    \author[Leipzig University]{Christin Lotz%
  %
  \fnref{4}}
   \ead{christin.lotz@uni-leipzig.de} 
    \author[Leipzig University]{Anne Deiglmayr%
  %
  \fnref{5}}
   \ead{anne.deiglmayr@uni-leipzig.de} 
      \affiliation[Leipzig University]{
    organization={Division of Empirical School and Classroom Research,
Institute of Educational Sciences},addressline={Marschnerstr.
29},city={Leipzig},postcode={4109},state={Sachsen},country={Germany},}
    \affiliation[Max Planck Institute for Evolutionary Anthropology]{
    organization={Department},addressline={Deutscher Pl.
6},city={Leipzig},postcode={4103},state={Sachsen},country={Germany},}
    \affiliation[University of Greifswald]{
    organization={Division of Health and Prevention, Institute of
Psychology},addressline={Robert-Blum-Str.
13},city={Greifswald},postcode={17489},state={Mecklenburg-Vorpommern},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{Conceptualization, Methodology, Formal analysis,
Investigation, Data Curation, Writing - Original Draft}
    \fntext[2]{Conceptualization, Methodology, Writing - Review \&
Editing, Supervision}
    \fntext[3]{Formal analysis, Data Curation, Writing - Review \&
Editing}
    \fntext[4]{Conceptualization, Methodology, Writing - Review \&
Editing}
    \fntext[5]{Conzeptualization, Methodology, Writing - Review \&
Editing, Supervision}

@@

Fig.\ref{fig.1} shows, in a simplified way, how classroom events affect
teachers' stress level, according to the adaptation of the Lazarus model
to teacher stress proposed by \citet{van2006stress}: When potential
stressors (e.g., classroom disruptions) occur during teaching (1),
teachers intuitively judge how disruptive the event is in a primary
appraisal (2). If potential stressors are judged as threatening, i.e.,
as actual stressors (3), teachers consider whether they have sufficient
resources for coping with the stressors (4). Teachers utilize these
resources in trying to cope with the stressors, e.g., by employing
classroom management strategies (5). In cases when coping fails, stress
ensues, often accompanied by physiological reactions like increased HR
(6). As part of the coping process, and dependent on its outcomes,
teachers re-evaluate the stressor (7).

As shown in Fig.\ref{fig.1}, teachers' primary and secondary appraisal
as well as coping attempts are influenced by professional experience. As
professional experience grows, teachers develop cognitive scripts for
managing classroom events, resulting in more complex and problem-focused
classroom management skills \citep{wolff2021classroom}, and thus more
effective coping and less stress. Empirically, classroom management
skills and problem-focused coping styles are linked to fewer instances
of emotional exhaustion \citep{maslach2001job, clunies2008self}. Novices
in the teaching profession, on the other hand, face considerable stress
and often feel overwhelmed by the demands of teaching
\citep{ophardt2017klassenmanagement, wolff2015keeping, klusmann2012berufliche}
with many leaving the profession within the first five years
\citep{ingersoll2003}. Accordingly, when resources are lacking and
coping fails, negative consequences for health (e.g., burnout) and for
work (e.g., high turnover rates) can arise
\citep{jalongo2006, unterbrink2007, aloe2014multivariate}, highlighting
the importance of professional expertise for managing teacher stress
\citep{fisher2011}.

\captionsetup[figure]{
    labelsep=colon,      % Period after Figure label
    font=footnotesize,           % Smaller font size for APA style
    justification=justified, % Left-justified captions, as per APA guidelines
    singlelinecheck=false % Avoid centering if the caption is single-line
}

\captionsetup[table]{
    labelsep=colon,      % Period after Figure label
    font=footnotesize,           % Smaller font size for APA style
    justification=justified, % Left-justified captions, as per APA guidelines
    singlelinecheck=false % Avoid centering if the caption is single-line
}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1\textwidth]{images/Model_Teacher_Stress_adapted_new.pdf}
  \caption{A model of teacher stress (adapted from van Dick 2006, p.37, modified by the authors).}
  \label{fig.1}
\end{figure}

@@

In addition, we selected five representative 10-minute intervals from
the five phases of the lab study (see Fig.\ref{fig.2}) in order to test
the hypotheses that, regarding HR levels, teachers' HR would be the
highest during the micro-teaching unit, compared to all other phases
(Hypothesis 1a), and, regarding HR slopes, that teachers' HR would
increase while they were preparing for teaching (pre-teaching interval),
but decrease in all of the following intervals, i.e.~when they were
habituating to and recovering from the stressful micro-teaching unit
(Hypothesis 1b).

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  We further explored whether teaching experience made a difference in
  how teachers' HR reacted to the classroom disruptions. We expected
  more experienced teachers to be less stressed by the classroom events
  (Hypothesis 2a). In addition, we examined the relations between
  teachers' subjective appraisals of the classroom events (specifically,
  the disruptiveness of the events, and their confidence in dealing with
  them) and teachers' HR level, beyond the explanatory power of teaching
  experience. We expected higher HR levels for teachers who felt more
  disrupted, regardless of their teaching experience (Hypothesis 2b),
  and lower HR levels for teachers who felt more confident, regardless
  of teaching experience (Hypothesis 2c). We hypothesized that each of
  the three predictors (\emph{teaching experience, disruption appraisal,
  confidence appraisal}) uniquely contributed to explaining variance in
  teachers' HR levels (Hypothesis 2d). In addition, we exploratively ran
  analogous analyses for the \emph{changes} in HR (i.e., slopes).
\end{enumerate}

@@

Each participant came to the lab for a period of approximately two hours
in total, and each participant underwent the same phases (see
Fig.\ref{fig.2}): In the \emph{pre-teaching phase}, the experimenter
welcomed the participants and helped them put on the fitness tracker.
This was followed by a warm-up session to familiarize the participants
with the laboratory setting and the class. This phase took about 10-15
minutes and participants spent this time mostly standing or slowly
walking around. During the \emph{teaching phase}, the participants held
their self-prepared micro-teaching unit to a class of three trained
actors who performed nine, potentially disruptive, classroom events
(e.g., chatting with a neighbor, heckling, looking at the phone; see
Tab.\ref{tab_a1} in the supplementary material for an overview and
categorization of all events; and Fig.\ref{fig.a2} in the supplementary
material for a depiction of the laboratory setting of the micro-teaching
unit). The topic and class level of the teaching unit could be freely
chosen by the teachers with the only requirement that the unit had to be
an introductory lesson, and had to consist of supervised individual work
and / or frontal teaching. The micro-teaching unit lasted about 15-20
minutes. Participants spent this time mostly standing or slowly walking
around. While teaching, participants wore eye-tracking glasses, and
their lesson was video-recorded. After having completed the
micro-teaching unit, in the \emph{post-teaching phase}, participants
filled in questionnaires for approximately 10-15 minutes: a brief
computer-based survey of sociodemographic data (e.g., teaching
experience, gender, studied school type, studied school subjects,
extracurricular teaching activities), and a short knowledge test that
was irrelevant to the present study. In the \emph{interview phase},
participants engaged in a Stimulated Recall Interview (SRI). During the
SRI, participants sat in front of a computer monitor and watched the
video of their own lesson from the ego perspective, as recorded through
the eye-tracking glasses. The experimenter stopped the video each time
one of the nine classroom events happened, and asked five open-ended,
and three rating questions per event. Two of the rating questions are
relevant to the present study: the disruption and the confidence
appraisal ratings (see Measures). The interview lasted about 45-60
minutes. Finally, in the \emph{end phase}, participants filled in
another questionnaire irrelevant to the present study, which lasted
about 10-15 minutes. \newpage

\begin{wrapfigure}[36]{r}{0.5\textwidth}
  \centering
  \includegraphics[width=0.5\textwidth]{images/Timeline_smaller.pdf}
  \caption{Procedure of the two-hour-long study consisting of five phases with five representative 10-minute intervals.}
  \label{fig.2}
\end{wrapfigure}

@@

Since we aimed to keep measurement intervals comparable between study
phases, we aggregated HR over a representative 10-minute interval within
each phase (cf.~Fig.\ref{fig.2}). Previous research has indicated that
10-minute intervals are a useful duration for analyzing PPG data
\citep{lu2008can}. The intervals were selected based on the following
rules: The \emph{pre-teaching interval} (\(I_1\)) comprised the first 10
minutes after the fitness tracker had been put on. The \emph{teaching
interval} (\(I_2\)) started two minutes after the lesson had started.
This interval was of the highest relevance to our study. We explicitly
chose an early 10-minute interval within the teaching phase, as previous
studies revealed that the beginning of a lesson is most demanding and
potentially stressful with regards to teacher-student interaction
\citep{donker2018, claessens2017positive}. The \emph{post-teaching
interval} (\(I_3\)) started immediately after the end of the teaching
unit. The \emph{interview interval} (\(I_4\)) was defined as the mid-10
minutes between the end of the teaching unit and the time point when the
fitness tracker was taken off. All participants were being interviewed
during this interval. The end interval (\(I_5\)) comprised the last 10
minutes before the fitness tracker was taken off.

@@

To enable visual inspection of HR trends, we displayed smoothed teacher
HR over the course of the recording.\footnote{The curve was smoothed
  using the geom\_smooth() function from the ggplot2 package in R
  \citep{ggplot2} based on the smoothing method LOESS (Locally Estimated
  Scatterplot Smoothing). This method fits a polynomial surface
  determined by one or more numerical predictors, using local fitting.}
We visually compared unstandardized and standardized HR trends over the
two-hour recording period.\footnote{Note that the study exceeded the
  planned duration of two hours for a few participants. To avoid
  distortions when mapping the HR over the course of the study (see
  Fig.\ref{fig.3}), the endpoint was set at two hours for all
  participants, even though data from later time points was used in the
  \emph{end interval} for a few participants.} For all further analyses,
we used standardized rather than unstandardized HR values.

@@

For testing Hypothesis 1b, concerning HR changes within each interval,
we first conducted a linear estimation of the increase or decrease in
standardized HR values over time for each participant. To this end, we
used fixed intercept fixed slope regression models
\citep{gelman2006data} for each interval to estimate intercepts and
linear slopes for each individual, which were then averaged across
individuals.\footnote{Although this procedure does not account for
  nonmonotonic progressions in individual HR, a graphical evaluation
  revealed that the linear estimates corresponded well to the majority
  of the cases (see Fig.\ref{fig.a3} to \ref{fig.a7} in the
  supplementary material).} We tested Hypothesis 1b based on the
unstandardized estimates of mean slopes (one estimate per participant
per interval).

@@

Means, standard deviations, and range of teachers' unstandardized and
standardized HR for the entire study period, and for the five intervals,
are shown in Tab.\ref{tab_1}. Fig.\ref{fig.3} displays the
unstandardized and standardized HR trends, respectively, over the course
of the entire study period. HR initially increased, peaked, and then
decreased, with the unstandardized and standardized HR graphs showing
high similarity. Thus, for all further analyses, we used participants'
standardized HR values.

Fig.\ref{fig.4} shows the distribution of teachers' mean standardized HR
for the five intervals. Repeated measures ANOVA revealed significant
differences in mean standardized HR between intervals,
\(F(4, 400) = 260.62\), \(p < .05\), \(d = 1.60\) (large effect).
Planned contrasts indicated that, as hypothesized (Hypothesis 1a), mean
standardized HR was significantly higher in the \emph{teaching interval}
(\(I_2\)) than in all other intervals, specifically, the
\emph{pre-teaching interval} (\(I_1\); \(t(400) = -10.08\), \(p < .05\),
\(d = 1.034\); large effect), the post-teaching interval (\(I_3\);
\(t(400) = -6.94\), \(p < .05\), \(d = 1.37\); large effect), the
interview interval (\(I_4\); \(t(400) = 15.00\), \(p < .05\),
\(d = 3.29\); large effect), and the end interval (\(I_5\));
\(t(400) = 22.54\), \(p < .05\), \(d = 4.64\); large effect).

Next, we examined HR changes (i.e., mean slopes) within each interval to
test the hypothesis that HR increased during the \emph{pre-teaching
phase} and decreased during all other phases (Hypothesis 1b). The mean
intercepts and mean slopes, complemented by their standard deviations
for each interval, are shown in Tab.\ref{tab_2}. The mean slope of the
\emph{pre-teaching interval} (\(I_1\)) was significantly positive,
indicating an increase in HR, as hypothesized. Further, the mean slopes
of the \emph{teaching interval} (\(I_2\)), \emph{post-teaching interval}
(\(I_3\)), and \emph{interview interval} (\(I_4\)) were significantly
negative, indicating a decrease in HR. For the last interval, the
\emph{end interval} (\(I_5\)), the mean slope was negative, but did not
differ significantly from zero.

@@

@@

\begin{table}[ht]
    \centering
    \begin{tabularx}{\textwidth}{lXXXXXX}
        \toprule
        Interval  & \multicolumn{2}{c}{\textit{M} (\textit{SD})} & \multicolumn{2}{c}{$p$} \\ & Intercept & Slope & Intercept & Slope \\
        \midrule
        ($I_1$) Pre-teaching interval & 0.052 (0.820) & 0.085 (0.133) & .57 & $< .05$ \\
        ($I_2$) Teaching interval & 1.025 (0.690) & -0.039 (0.108) & $< .05$ & $< .05$ \\
        ($I_3$) Post-teaching interval & 0.549 (0.547) & -0.060 (0.101) & $< .05$ & $< .05$ \\
        ($I_4$) Interview interval & -0.617 (0.614) & -0.022 (0.070) & $< .05$ & $< .05$ \\
        ($I_5$) End interval & -1.004 (0.500) & -0.012 (0.074) & $< .05$ & .14 \\
        \bottomrule
    \end{tabularx}
    \caption{Analysis (\textit{M, SD, p}-values) for the mean intercepts and the mean slopes for the five intervals.}
    \label{tab_2}
\end{table}

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{plots_publication/loess_plot_std_unstd_new.pdf}
  \caption{Overall course of the HR with the unstandardized HR in BPM shown in Fig.3a. and the z-standardized HR shown in Fig.3b. for the planned 2-hour study.}
  \label{fig.3}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=1\textwidth]{plots_publication/box_plot.pdf}
  \caption{Distribution of the standardized heart rate means in the five intervals.}
  \label{fig.4}
\end{figure}

@@

Tab.\ref{tab_3} shows the raw correlations among mean standardized
HR/mean slopes (see Tab.\ref{tab_2} for means and standard deviations),
teaching experience (\(M = 5.64\), \(SD = 9.46\)), disruption appraisal
(\(M = 5.19\), \(SD = 2.87\)), and confidence appraisal (\(M = 7.81\),
\(SD = 1.97\)). With a few notable exceptions, correlations with HR
measures were mostly very small and statistically non-significant
(Tab.\ref{tab_3}). Correlations between teaching experience and
appraisals (not shown in Tab.\ref{tab_3}) were substantial: more
experienced teachers gave lower disruption appraisals (\(r = -.36\)),
and higher confidence appraisals (\(r = .44\)). Moreover, the two
appraisal variables were negatively correlated (\(r = -.37\)).

Tab.\ref{tab_4} shows the results of the regression analyses. Teaching
experience significantly predicted mean standardized HR only in the
\emph{interview interval} (Tab.\ref{tab_4}, \emph{interview interval},
Model 1), indicating a higher mean standardized HR for teachers with
more teaching experience. This relationship is, in fact, in the opposite
direction as predicted by Hypothesis 2a. Neither adding disruption
appraisal (Hypothesis 2b) nor adding confidence appraisal (Hypothesis
2c) increased the amount of explained variance to a statistically
significant extent.

When considering the effects of the three predictors in concert
(Hypothesis 2d), mean standardized HR was significantly predicted only
by disruption appraisal, and only in the \emph{post-teaching interval}
(Tab.\ref{tab_4}, \emph{post-teaching interval}, Model 4), indicating a
higher mean standardized HR for teachers who felt more disrupted by the
classroom events, when controlling for the other variables.

Concerning the explorative investigation of the effects of teaching
experience and subjective appraisals on \emph{changes} (i.e., mean
slopes) in teachers' HR, teaching experience significantly predicted the
mean slope in the \emph{pre-teaching interval} (Tab.\ref{tab_4},
Pre-teaching interval, Model 1), indicating a less steep HR increase in
teachers with more teaching experience. For all other intervals, no
variable had significant predictive value.

@@
